---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "80%",
  dev = "ragg_png",
  dpi = 90
)
```

# gguidance

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/gguidance)](https://CRAN.R-project.org/package=gguidance)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/teunbrand/gguidance/workflows/R-CMD-check/badge.svg)](https://github.com/teunbrand/gguidance/actions)
[![Codecov test coverage](https://codecov.io/gh/teunbrand/gguidance/branch/master/graph/badge.svg)](https://app.codecov.io/gh/teunbrand/gguidance?branch=master)
<!-- badges: end -->

The goal of gguidance is to provide additional guides to the ggplot2 ecosystem. Guides are re-implemented within the ggproto system to make use of the flexible extension framework.

Please note that this repo is still being worked on and, while probably usable, isn't finished.

## Installation

You can install the development version of gguidance from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("teunbrand/gguidance")
```

## Example

Here is a simple example of how to use an axis guide and a colour bar guide.

```{r example}
library(gguidance)

ggplot(msleep, aes(bodywt, sleep_total)) +
  geom_point(aes(colour = sleep_rem)) +
  scale_colour_viridis_c(guide = "colourbar_cap") +
  scale_x_log10(guide = "axis_log")
```

## Progress

So far, the following has been implemented:

* **Axis guides**
   * `guide_axis_vanilla()`: a re-implementation of `guide_axis()`.
   * `guide_axis_ext()`: Extended options for axes.
   * `guide_axis_minor()`: Axes with minor ticks.
   * `guide_axis_trunc()`: Axes with truncated axis lines.
   * `guide_axis_log()`: Axes with log10-based tickmarks.
* **Legend guides**
   * `guide_legend_vanilla()`: a re-implementation of `guide_legend()`.
* **Colour bar guides**
   * `guide_colourbar_vanilla()`: a re-implementation of `guide_colourbar()`.
   * `guide_colourbar_cap()`: With cap decorations at the ends.

## Notes

A medium-term goal is to move and refactor guides from the {ggh4x} package to {gguidance}. A longer-term goal is to make this package a hub for additional axes, legends and other guides.
