---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "80%",
  dev = "ragg_png",
  dpi = 90
)
```

# gguidance

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/gguidance)](https://CRAN.R-project.org/package=gguidance)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/teunbrand/gguidance/workflows/R-CMD-check/badge.svg)](https://github.com/teunbrand/gguidance/actions)
[![Codecov test coverage](https://codecov.io/gh/teunbrand/gguidance/branch/master/graph/badge.svg)](https://app.codecov.io/gh/teunbrand/gguidance?branch=master)
<!-- badges: end -->

> **Warning**
> You're looking at an experimental branch that explores guide extensions with an upcoming ggproto overhaul in ggplot2.

The goal of gguidance is to provide additional guides to the ggplot2 ecosystem. 

Please note that this repo is still being worked on and, while probably usable, isn't finished.

## Installation

You can install the development version of gguidance from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("tidyverse/ggplot2", ref = remotes::github_pull("5096"))
remotes::install_github("teunbrand/gguidance", ref = "main")
```

## Examples

Demonstrating a 'cross legend'.

```{r example}
library(gguidance)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = paste(cyl, year))) +
  guides(colour = "legend_cross")
```
