---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "80%",
  dev = "ragg_png",
  dpi = 90
)
```

# gguidance

<!-- badges: start -->
<!-- badges: end -->

The goal of gguidance is to provide additional guides to the ggplot2 ecosystem. Guides are re-implemented within the ggproto system to make use of the flexible extension framework.

## Installation

You can install the development version of gguidance from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("teunbrand/gguidance")
```

## Example

As of now, the only thing of note is `guide_colourbar_cap()`.

```{r example}
library(gguidance)

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = cty)) +
  scale_colour_viridis_c(guide = "colourbar_cap")
```

## Progress

So far, the following has been implemented:

* **Axis guides**
   * `guide_axis_vanilla()`: a re-implementation of `guide_axis()`.
* **Legend guides**
   * `guide_legend_vanilla()`: a re-implementation of `guide_legend()`.
* **Colour bar guides**
   * `guide_colourbar_vanilla()`: a re-implementation of `guide_colourbar()`.
   * `guide_colourbar_cap()`: With cap decorations at the ends.

## Notes

A medium-term goal is to move and refactor guides from the {ggh4x} package to {gguidance}. A longer-term goal is to make this package a hub for additional axes, legends and other guides.
