---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "80%",
  dev = "ragg_png",
  dpi = 90
)
```

# gguidance

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/gguidance)](https://CRAN.R-project.org/package=gguidance)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![Codecov test coverage](https://codecov.io/gh/teunbrand/gguidance/branch/master/graph/badge.svg)](https://app.codecov.io/gh/teunbrand/gguidance?branch=master)
[![R-CMD-check](https://github.com/teunbrand/gguidance/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/teunbrand/gguidance/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

> **Warning**
> You're looking at an experimental branch that explores guide extensions with an upcoming ggproto overhaul in ggplot2.

The goal of gguidance is to provide additional guides to the ggplot2 ecosystem. 

Please note that this repo is still being worked on and, while probably usable, isn't finished.

## Installation

You can install the development version of gguidance from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("tidyverse/ggplot2", ref = remotes::github_pull("5096"))
remotes::install_github("teunbrand/gguidance", ref = "main")
```

## Examples

Let's first set up a basic plot to experiment with

```{r setup}
library(gguidance)

p <- ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  labs(
    x = "Engine displacement",
    y = "Highway miles per gallon"
  ) +
  theme(axis.line = element_line())
```

### Legends

A 'cross legend' can show two variables in a single legend.

```{r legend_cross}
p + aes(colour = paste(cyl, year)) +
  guides(colour = "legend_cross")
```

### Colour bars

A capped colour bar:

```{r colourbar_cap}
p + aes(colour = cty) +
  scale_colour_viridis_c(guide = "colourbar_cap")
```

Using a violin as a colour guide:

```{r colour_violin}
p + aes(colour = cty) +
  scale_colour_viridis_c(guide = guide_colour_violin(density = mpg$cty))
```

### Axes

Using subtitles.

```{r axis_extend}
p + guides(x = guide_axis_extend(subtitle = c("Less", "More")))
```

Using minor ticks.

```{r axis_minor}
p + guides(x = "axis_minor")
```

Using capped lines.

```{r axis_cap}
p + guides(x = "axis_cap")
```

With bracketed ranges.

```{r axis_nested}
ggplot(mpg, aes(class, displ)) +
  geom_boxplot() +
  guides(x = guide_axis_nested(
    range_start = c(0.5, 3.5),
    range_end   = c(4.5, 6.5),
    range_name  = c("First range", "Second range"),
    bracket     = "square" 
  ))
```


Using a table as an axis guide.

```{r axis_table}
# Creating summary table
my_table <- lapply(split(mpg[, c("displ", "cty", "hwy")], mpg$cyl), colMeans)
my_table <- as.data.frame(do.call(rbind, my_table))
my_table[] <- lapply(my_table, scales::number, accuracy = 0.01)
my_table$cyl <- rownames(my_table)

# Use summary table as axis guide
ggplot(mpg, aes(factor(cyl), displ)) +
  geom_boxplot() +
  guides(x = guide_axis_table(table = my_table, key_col = cyl))
```
