---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gguidance

<!-- badges: start -->
<!-- badges: end -->

The goal of gguidance is to provide additional guides to the ggplot2 ecosystem. Guides are re-implemented within the ggproto system to make use of the flexible extension framework.

## Installation

You can install the development version of gguidance from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("teunbrand/gguidance")
```

## Example

As of now, there is only an re-implementation of `ggplot2::guide_axis()`.

```{r example}
library(gguidance)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  guides(x = guide_axis_vanilla())
```

## Notes

A medium-term goal is to move and refactor guides from the {ggh4x} package to {gguidance}. A longer-term goal is to make this package a hub for additional axes, legends and other guides.
