---
title: "Example Gallery"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dev = "ragg_png"
)
```

```{r setup}
library(legendry)
```

This article is intended to showcase a few strategies to use the building blocks that legendry provide to build more specialised guides.

## Rugs

An easy way to wrap a rug guide is to combine `primitive_ticks()` with a manual key. 
Axis guides cannot lookup values from the plot data, so we would need to feed the rug data manually.

```{r}
guide_rug <- function(x, ...) {
  primitive_ticks(key = key_manual(x), ...)
}

rug_wt <- guide_rug(mtcars$wt)
rug_mpg <- guide_rug(mtcars$mpg)
```

The simplest way to use the guide would be as secondary axis.

```{r}
base <- ggplot(mtcars, aes(wt, mpg, colour = drat)) +
  geom_point() +
  labs(
    x = "Weight (1000 lbs)",
    y = "Miles per gallon",
    colour = "Rear axle\nratio"
  )
base + guides(x.sec = rug_wt, y.sec = rug_mpg)
```

The guide can also easily stacked with typical axis guides. 
To style the ticks drawn by the rug differently than the regular axis ticks, you can give the rug guide a local theme.

```{r}
long_ticks <- theme_guide(
  ticks.length = unit(0.5, "cm"),
  ticks = element_line(colour = "dodgerblue")
)

base + guides(
  x = compose_stack("axis", rug_wt),
  y = compose_stack("axis", guide_rug(mtcars$mpg, theme = long_ticks))
)
```

For using a rug at colour bars, you can feed a rug to the `guide_colbar(second_guide)` argument.
Note that colour-bar guides by default have inward facing, white ticks, so you might like to style them to look more like regular ticks.

```{r}
rug_drat <- guide_rug(
  x = mtcars$drat,
  theme = theme_guide(
    ticks.length = unit(2.75, "pt"),
    ticks = element_line(colour = "black", linewidth = 0.5)
  )
)

base + guides(colour = guide_colbar(second_guide = rug_drat))
```

