---
title: "A guided tour"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gguidance)
library(scales)
```

This article walks through all full guide extensions to give a broad overview of the 'easy' way to make use of gguidance.
Rest assured, there are harder ways, but these will be covered in a separate article.

## Axes

Axes truly are the bread and butter of guides.
Naturally, axes shine brightest as guides for positions like `x` and `y` but can moonlight as auxiliary guides as well.

### Where (not) to apply

In gguidance, the staple axis is `guide_axis_custom()`.
At a first glance, these axes are utterly unremarkable and very much mirror `ggplot2::guide_axis()` by design.

```{r}
# Turn on axis lines
theme_update(axis.line = element_line())

# A standard plot
standard <- ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  labs(
    x = "Engine Displacement (Litres)",
    y = "Highway Miles per Gallon"
  )

standard + guides(
  x = "axis_custom",
  y = "axis_custom"
)
```

In terms of novelty, the only 'extra' option these axes offer is to display bidirectional tick marks.

```{r}
p <- standard + 
  scale_x_continuous(guide = guide_axis_custom(bidi = TRUE)) +
  scale_y_continuous(guide = guide_axis_custom(bidi = TRUE))
p
```

However, `guide_axis_custom()` is more flexible than `ggplot2::guide_axis()`.
In ggplot2, you'd typically have to switch to `ggplot2::guide_axis_theta()` to display an axis for the `theta` coordinate of a polar plot.
The custom axis knows how to fit into polar coordinates, so no such fuss is needed when switching to polar coordinates.

```{r}
p + coord_radial(start = 1.25 * pi, end = 2.75 * pi)
```

Argueably, the custom guide is a little bit too flexible. 
It has exactly no scruples in displaying arbitrary continuous aesthetics, like `colour` in the plot below.
As you can see, it is not informative as a colour guide and for this reason I'd advice against it.
Why this unadvised yet possible is a topic that resurfaces later in this article.

```{r}
standard + aes(colour = cty) +
  guides(colour = "axis_custom")
```

In summary, `guide_axis_custom()` is a flexible guide that can be used in any and all position aesthetic, and can (but should not) be used for other continuous aesthetics.

### Keys

The way guides exchange information with scales is through so-called 'keys', which you can retrieve using `ggplot2::get_guide_data()`.
In gguidance, you have to option to change what information is exchanged through the `key` argument.

If you've browsed the documentation of `guide_axis_custom()`, you might have noticed that there is no visible option to display minor ticks.
To extract the minor ticks, we can use `key = "minor"`, which invokes the `key_minor()` function to extract the minor breaks from the scale.

```{r}
standard +
  guides(x = guide_axis_custom(key = "minor"))
```

Likewise, there is no logarithmic variant of `guide_axis_custom()`, as this is again a specialty extraction operation.
The equivalent of using `ggplot2::guide_axis_logticks()` is to use `key = "log"`.
If you need to tweak the options you can use the `key_log()` function instead of the keyword, as is the case for the y-axis in the plot below where we manually log-transformed the `brainwt` variable.

```{r}
ggplot(msleep, aes(bodywt, log2(brainwt))) +
  geom_point(na.rm = TRUE) +
  scale_x_log10(
    name = "Body weight (kg)",
    guide = guide_axis_custom(key = "log")
  ) +
  scale_y_continuous(
    name = "Brain weight (kg)",
    guide = guide_axis_custom(key = key_log(prescale_base = 2))
  ) +
  theme(axis.ticks.length = unit(2, "mm"))
```

Alternatively, you can construct a key yourself by using `key_manual()`, where you have to provide the value for the aesthetic, and the label you want to use.

```{r}
freq <- table(mpg$class) |> as.data.frame()
key_manual(aesthetic = freq$Var1, label = paste0("n = ", freq$Freq))
```

The non-standard evaluation equivalent of `key_manual()` is `key_map()` which works in a similar way, but is more suited when piping data frames.
A nice thing about these function is that they are a good way to generate secondary axes for discrete scales.
At the time of writing, this isn't really possible in ggplot2 itself.

```{r}
key <- table(mpg$class) |>
  as.data.frame() |>
  key_map(aesthetic = Var1, label = paste0("n = ", Freq))

ggplot(mpg, aes(class, displ)) +
  geom_boxplot() +
  labs(
    x = "Type of car",
    y = "Engine Displacement (Litres)"
  ) +
  guides(x.sec = guide_axis_custom(key = key))
```

### Novelty axes

First have to implement some.

## Colours

The `colour` and `fill` aesthetics are wonderful to build guides for, as they can apply to pretty much anything.
First, we'll take a gander at some variants of colour bars before we gander at rings.

### Bars and steps

Two variants for colour guides exist in {gguidance}:

1. `guide_colourbar_custom()` that reflects `guide_colourbar()`
2. `guide_coloursteps_custom()` that reflects `guide_coloursteps()`.

When used in a standard fashion, they look very similar to their vanilla counterparts.

```{r}
#| fig-show: hold
standard <- standard + 
  aes(colour = cty) +
  labs(colour = "City Miles\nper Gallon")

standard +
  scale_colour_viridis_c(guide = "colourbar_custom") +
  labs(title = "Custom colour bar")

standard +
  scale_colour_viridis_b(guide = "coloursteps_custom") +
  labs(title = "Custom colour steps")
```

Please not that the following paragraphs apply equally to `guide_coloursteps_custom()`, but we'll take `guide_colourbar_custom()` for examples.

#### Caps

The thing that sets these guides apart is that they have indicators for when the data goes out-of-bounds.
The most common case where you have out-of-bounds data, is when you set the scale limits to be narrower than the data range.
In the plot below, the `cty` variable has a few observation below the lower limit of 10, and a few above the upper limit of 30.
Typically, these are displayed in the `na.value = "grey"` colour.
The bars display that these data are out-of-bounds by the gray 'caps' at the two ends of the bar.

```{r, dev = 'png', dev.args=list(type = 'cairo')}
standard +
  scale_colour_viridis_c(
    limits = c(10, 30),
    guide = "colourbar_custom"
  )
```

You can change the out-of-bounds strategy, the `oob` argument of the scale, to have the caps reflect the colour that out-of-bounds data has acquired.

```{r}
standard +
  scale_colour_viridis_c(
    limits = c(10, 30), oob = oob_squish,
    guide = "colourbar_custom"
  )
```

You can also force the caps to appear, even when there are no out-of-bounds data, or force the cap colour to be consistent with the scale.

```{r}
standard +
  scale_colour_viridis_c(
    guide = guide_colourbar_custom(
      show = c(FALSE, TRUE), 
      oob = "squish"
    )
  )
```

The shape of the cap needn't be a triangle. You can set the shape to any of the built-in cap shapes.

```{r}
standard +
  scale_colour_viridis_c(
    guide = guide_colourbar_custom(
      show = TRUE, oob = "squish",
      shape = "arch"
    )
  )
```

The caps can be provided as a string naming a cap function, like `"arch"` that invokes `cap_arch()`.
Below follows an overview of all the build-in cap shapes.

```{r}
caps <- list(
  none = cap_none(),
  triangle = cap_triangle(),
  round = cap_round(),
  arch = cap_arch(),
  ogee = cap_ogee()
)

caps <- cbind(
  as.data.frame(do.call(rbind, caps)),
  shape = factor(rep(names(caps), lengths(caps) / 2), names(caps))
)

ggplot(caps, aes(x, y)) +
  geom_path(arrow = arrow()) +
  facet_wrap(~ shape) +
  coord_equal()
```

It is most certainly possible to use shapes of your own imagination as well.
To provide your own shape, use a numeric matrix that:

* Has 2 columns corresponding to the x and y coordinates.
* Has at least 2 rows.
* Only has positive values for the 2nd column (y).
* Start at the (0, 0) coordinate.
* End at the (1, 0) coordinate.

You can see in the shapes above that these requirements all hold for the built-in shapes.
Such a matrix can be given to the `shape` argument of the guide.

```{r}
hourglass_cap <- cbind(
  x = c(0, 1, 0, 1),
  y = c(0, 1, 1, 0)
)

standard +
  scale_colour_viridis_c(
    guide = guide_colourbar_custom(
      show = TRUE, oob = "squish",
      shape = hourglass_cap
    )
  )
```

#### Side-guides

The colour bars come with a small party trick: the two rows of tick marks are separate axes masquerading as parts of the colour bar.
It becomes easier to see once you wash away their make-up with `vanilla = FALSE`.

```{r}
standard +
  scale_colour_viridis_c(
    guide = guide_colourbar_custom(vanilla = FALSE)
  )
```

This trick allows you to tailor the colour bar to your liking on separate sides.
You can use this to invoke any of the tricks described in the axis section, like setting minor ticks, or swap out axes for an annotation-primitive like `primitive_bracket()`.

```{r}
brackets <- 
  key_range_manual(
    start = c(9, 25),
    end   = c(19, 30), 
    name  = c("A", "B")
  ) |>
  primitive_bracket(bracket = "square")

standard +
  scale_colour_viridis_c(
    minor_breaks = breaks_width(1),
    guide = guide_colourbar_custom(
      first_guide = guide_axis_custom("minor"),
      second_guide = brackets
    )
  )
```

### Rings

## Legends

Well they'd have to be implemented first before there is something to tell.

To be continued once I find more time writing about this.
