---
title: "A guided tour"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(gguidance)
library(scales)
```

This article walks through all full guide extensions to give a broad overview of the 'easy' way to make use of gguidance.
Rest assured, there are harder ways, but these will be covered in a separate article.

## Axes

Axes truly are the bread and butter of guides.
Naturally, axes shine brightest as guides for positions like `x` and `y` but can moonlight as auxiliary guides as well.

### Where (not) to apply

In gguidance, the staple axis is `guide_axis_base()`.
At a first glance, these axes are utterly unremarkable and very much mirror `ggplot2::guide_axis()` by design.

```{r}
# Turn on axis lines
theme_update(axis.line = element_line())

# A standard plot
standard <- ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  labs(
    x = "Engine Displacement (Litres)",
    y = "Highway Miles per Gallon"
  )

standard + guides(
  x = "axis_base",
  y = "axis_base"
)
```

In terms of novelty, the only 'extra' option these axes offer is to display bidirectional tick marks.

```{r}
p <- standard + 
  scale_x_continuous(guide = guide_axis_base(bidi = TRUE)) +
  scale_y_continuous(guide = guide_axis_base(bidi = TRUE))
p
```

However, `guide_axis_base()` is more flexible than `ggplot2::guide_axis()`.
In ggplot2, you'd typically have to switch to `ggplot2::guide_axis_theta()` to display an axis for the `theta` coordinate of a polar plot.
The custom axis knows how to fit into polar coordinates, so no such fuss is needed when switching to polar coordinates.

```{r}
p + coord_radial(start = 1.25 * pi, end = 2.75 * pi)
```

Argueably, the custom guide is a little bit too flexible. 
It has exactly no scruples in displaying arbitrary continuous aesthetics, like `colour` in the plot below.
As you can see, it is not informative as a colour guide and for this reason I'd advice against it.
Why this unadvised yet possible is a topic that resurfaces later in this article.

```{r}
standard + aes(colour = cty) +
  guides(colour = "axis_base")
```

In summary, `guide_axis_base()` is a flexible guide that can be used in any and all position aesthetic, and can (but should not) be used for other continuous aesthetics.

### Nested axes

Currently, there is exactly 1 'novelty' axis and that is `guide_axis_nested()`.
Let's suppose we have 'nested' data, which for our purposes just means that discrete variables have some kind of categories or interactions to them that can be laid out in a nested fashion. 
A category for categories, if you will.

In the example below we have three super-categories 'Drink', 'Fruit' and 'Vehicle' which will have more granular categories like 'Coffee' and 'Pear' that belong to the super-categories.
We can use the `interaction()` function to paste together the name of the inner category with the name of the outer category.

```{r}
df <- data.frame(
  item = c("Coffee", "Tea", "Apple", "Pear", "Car"),
  type = c("Drink", "Drink", "Fruit", "Fruit", "Vehicle"),
  amount = c(5, 1, 2, 3, 1)
)

plain <- ggplot(df, aes(interaction(item, type), amount)) +
  geom_col()
plain + guides(x = "axis_nested")
```

Instead of just relying on formatting the labels correctly for splitting, you can also manually annotate the outer categories. To do this, we can use the `key_range_manual()` function to constructs the brackets as we see fit.

```{r}
my_key <- key_range_manual(
  start = c("Coffee", "Apple"),
  end   = c("Tea", "Pear"),
  name  = c("Drinks", "Fruits"), 
  level = 1
)

ggplot(df, aes(item, amount)) +
  geom_col() +
  scale_x_discrete(
    limits = df$item,
    guide = guide_axis_nested(
      regular_key = "auto",
      key = my_key
    )
  )
```

#### Brackets

To change the style of the range indicators, you can choose a different `bracket` setting.
The theme elements `gguidance.bracket` and `gguidance.bracket.size` control the styling and size of the line.
These settings have a shortcut in `theme_guide()`.

```{r}
plain + guides(x = guide_axis_nested(bracket = "curvy")) +
  theme_guide(
    bracket = element_line(colour = "blue"),
    bracket.size = unit(3, "mm")
  )
```

The brackets can be provided as a string naming a bracket function, like `"curvy"` that invokes `bracket_curvy()`.
Below follows an overview of all the build-in bracket shapes.

```{r}
brackets <- list(
  "atan"    = bracket_atan(),
  "chevron" = bracket_chevron(),
  "curvy"   = bracket_curvy(),
  "line"    = bracket_line(), 
  "round"   = bracket_round(),
  "sigmoid" = bracket_sigmoid(),
  "square"  = bracket_square()
)

brackets <- cbind(
  as.data.frame(do.call(rbind, brackets)),
  shape = factor(rep(names(brackets), lengths(brackets) / 2), names(brackets))
)

ggplot(brackets, aes(x, y)) +
  geom_path() +
  facet_wrap(~ shape) +
  coord_equal()
```

Quite possibly, there might be bracket shapes you want to use, but aren't built into gguidance.
Luckily, we can build custom brackets, using a numeric matrix that:

* Has 2 columns corresponding to the x and y coordinates.
* Has at least 2 rows.
* Only has values between 0 and 1.

The x-coordinate will be stretched along the axis, whereas y will be squished to fit the `gguidance.bracket.size` theme setting.
A custom bracket can just be provided to the `bracket` argument.

```{r}
zigzag <- cbind(
  x = seq(0, 1, length.out = 20),
  y = rep(c(0, 1), length.out = 20)
)

plain + guides(x = guide_axis_nested(bracket = zigzag))
```

#### Boxes

Alternatively, it is also possible to forego brackets altogether and use boxes instead.

```{r}
plain + guides(x = guide_axis_nested(type = "box"))
```

#### Customising

You needn't strictly use `guide_axis_nested()` with discrete data: you can use it with continuous data as well.
However, you'd need to provide a manual ranged key, such as one created by `key_range_manual()`/`key_range_map()`.

```{r}
presidents <- key_range_map(presidential, start = start, end = end, name = name)

eco <- ggplot(economics, aes(date, unemploy)) +
  geom_line() +
  labs(y = "Unemployment")
  
eco + guides(x = guide_axis_nested(key = presidents))
```

To customise the different depths of the bracketed text, you can give a list of text elements to the `levels_text` argument.

```{r}
presidents$.level <- rep(1:3, length.out = nrow(presidents))

eco + guides(x = guide_axis_nested(
  key = presidents,
  levels_text = list(
    element_text(face = "bold"),
    NULL,
    element_text(face = "italic")
  )
))
```

Alternatively, you can tailor many of the usual text formatting options by encoding these in the key.

```{r}
presidents <- key_range_map(
  presidential, 
  start = start, end = end, name = name,
  level = rep(1:4, length.out = nrow(presidential)),
  colour = ifelse(party == "Republican", "tomato", "dodgerblue")
)

eco + guides(x = guide_axis_nested(key = presidents))
```


## Colours

The `colour` and `fill` aesthetics are wonderful to build guides for, as they can apply to pretty much anything.
First, we'll take a gander at some variants of colour bars before we gander at rings.

### Bars and steps

Two variants for colour guides exist in {gguidance}:

1. `guide_colbar()` that reflects `guide_colourbar()`
2. `guide_colsteps()` that reflects `guide_coloursteps()`.

When used in a standard fashion, they look very similar to their vanilla counterparts.

```{r}
#| fig-show: hold
standard <- standard + 
  aes(colour = cty) +
  labs(colour = "City Miles\nper Gallon")

standard +
  scale_colour_viridis_c(guide = "colbar") +
  labs(title = "Custom colour bar")

standard +
  scale_colour_viridis_b(guide = "colsteps") +
  labs(title = "Custom colour steps")
```

Please note that the following paragraphs apply equally to `guide_colsteps()`, but we'll take `guide_colbar()` for examples.

#### Caps

The thing that sets these guides apart is that they have indicators for when the data goes out-of-bounds.
The most common case where you have out-of-bounds data, is when you set the scale limits to be narrower than the data range.
In the plot below, the `cty` variable has a few observation below the lower limit of 10, and a few above the upper limit of 30.
Typically, these are displayed in the `na.value = "grey"` colour.
The bars display that these data are out-of-bounds by the gray 'caps' at the two ends of the bar.

```{r, dev = 'png', dev.args=list(type = 'cairo')}
standard +
  scale_colour_viridis_c(
    limits = c(10, 30),
    guide = "colbar"
  )
```

You can change the out-of-bounds strategy, the `oob` argument of the scale, to have the caps reflect the colour that out-of-bounds data has acquired.

```{r}
standard +
  scale_colour_viridis_c(
    limits = c(10, 30), oob = oob_squish,
    guide = "colbar"
  )
```

You can also force the caps to appear, even when there are no out-of-bounds data, or force the cap colour to be consistent with the scale.

```{r}
standard +
  scale_colour_viridis_c(
    guide = guide_colbar(
      show = c(FALSE, TRUE), 
      oob = "squish"
    )
  )
```

The shape of the cap needn't be a triangle. You can set the shape to any of the built-in cap shapes.

```{r}
standard +
  scale_colour_viridis_c(
    guide = guide_colbar(
      show = TRUE, oob = "squish",
      shape = "arch"
    )
  )
```

The caps can be provided as a string naming a cap function, like `"arch"` that invokes `cap_arch()`.
Below follows an overview of all the build-in cap shapes.

```{r}
caps <- list(
  none = cap_none(),
  triangle = cap_triangle(),
  round = cap_round(),
  arch = cap_arch(),
  ogee = cap_ogee()
)

caps <- cbind(
  as.data.frame(do.call(rbind, caps)),
  shape = factor(rep(names(caps), lengths(caps) / 2), names(caps))
)

ggplot(caps, aes(x, y)) +
  geom_path(arrow = arrow()) +
  facet_wrap(~ shape) +
  coord_equal()
```

It is most certainly possible to use shapes of your own imagination as well.
To provide your own shape, use a numeric matrix that:

* Has 2 columns corresponding to the x and y coordinates.
* Has at least 2 rows.
* Only has positive values for the 2nd column (y).
* Start at the (0, 0) coordinate.
* End at the (1, 0) coordinate.

You can see in the shapes above that these requirements all hold for the built-in shapes.
Such a matrix can be given to the `shape` argument of the guide.

```{r}
hourglass_cap <- cbind(
  x = c(0, 1, 0, 1),
  y = c(0, 1, 1, 0)
)

standard +
  scale_colour_viridis_c(
    guide = guide_colbar(
      show = TRUE, oob = "squish",
      shape = hourglass_cap
    )
  )
```

#### Side-guides

The colour bars come with a small party trick: the two rows of tick marks are separate axes masquerading as parts of the colour bar.
It becomes easier to see once you wash away their make-up with `vanilla = FALSE`.

```{r}
standard +
  scale_colour_viridis_c(
    guide = guide_colbar(vanilla = FALSE)
  )
```

This trick allows you to tailor the colour bar to your liking on separate sides.
You can use this to invoke any of the tricks described in the axis section, like setting minor ticks, or swap out axes for an annotation-primitive like `primitive_bracket()`.

```{r}
brackets <- 
  key_range_manual(
    start = c(9, 25),
    end   = c(19, 30), 
    name  = c("A", "B")
  ) |>
  primitive_bracket(bracket = "square")

standard +
  scale_colour_viridis_c(
    minor_breaks = breaks_width(1),
    guide = guide_colbar(
      first_guide = guide_axis_base("minor"),
      second_guide = brackets
    )
  )
```

### Rings

Aside from bars and steps, there is also an option to show the colour as a ring.
To understand why this might convenient, it can help to understand the type of data this is suitable for.
A prime example of cyclical data can be the month of the year.
The time between December and January is just one month, but when encoded numerically, the difference is 11 months.
This problem can show itself sometimes in periodic data, like housing sales below.

```{r}
housing <- 
  ggplot(
    subset(txhousing, city == "Houston"), 
    aes(date, volume, colour = month)
  ) +
  geom_line() +
  scale_y_continuous(
    name = "Total volume of sales",
    labels = dollar_format(scale = 1e-6, suffix = "M")
  ) +
  labs(
    x = "Date",
    colour = "Month"
  )

housing +
  scale_colour_viridis_c(limits = c(0, 12))
```

Every year we get a sharp colour transition in the winter.
The remedy for this problem is to use a cyclical palette.
The {scico} package offers some suitable cyclical palettes, like 'romaO', 'vikO', 'bamO', 'corkO' or 'brocO'.

```{r}
# Colours from scico::scico(12, palette = "romaO")
periodic_pal <- 
  c("#723957", "#843D3A", "#97552B", "#B08033", "#CBB45D", "#D5DA99", 
    "#B8DEC3", "#85C7CF", "#599FC4", "#4E73AB", "#5F4C81", "#723959")

housing + 
  scale_colour_gradientn(colours = periodic_pal, limits = c(0, 12))
```

This is already much better, but the guide itself does a poor job of displaying the cyclical nature of months.
To have this better reflected in the guide, you can use `guide_colring()`.

```{r}
housing + 
  scale_colour_gradientn(
    colours = periodic_pal, limits = c(1, 13),
    breaks = 1:12,
    guide = "colring"
  )
```

The 'thickness' of the donut can be controlled by the `legend.key.width` parameter, which by default is 1/5^th^ of the diameter. The outer diameter of the ring is controlled by the `legend.key.size` parameter, but multiplied by 5 for consistency with the colour bar multiplier.
Like custom colour bars, it is possible to set custom guides, but these are hoarded under the `inner_guide` and `outer_guide` to distinguish that they aren't first or second.

```{r}
housing + 
  scale_colour_gradientn(
    colours = periodic_pal, limits = c(1, 13),
    breaks = 1:12, minor_breaks = breaks_width(0.25),
    guide = guide_colring(
      outer_guide = guide_axis_base("minor"),
      inner_guide = "none"
    )
  ) +
  theme(
    legend.key.width = rel(2.5), # fill to center
    legend.key.size = unit(0.5, "cm") # actual size is 0.5 * 5 = 2.5 cm
  )
```

## Legends

Well they'd have to be implemented first before there is something to tell.

To be continued once I find more time writing about this.
